Edge sign prediction in general graphs and connection with Correlation Clustering

### 2017-07-03 assets/tikz/gtx_singleton_tikz.tex
minor wording and layout adjustments

    \draw[->] (0,-2) -- (0,-2.9);

    \node[vertex,fill=DodgerBlue,text=black] (s1) at (-2,-3.7) {$S_1^1$};
    \node[vertex,fill=DodgerBlue,text=black] (s2) at (2,-3.7) {$S_2^1$};
    \node[vertex,fill=DodgerBlue,text=black] (s3) at (0,-3.7) {$S_3^1$};
    \foreach \source/\dest in {3/2, 1/3}
    \draw[edge] (s\source) -- (s\dest);
  \end{tikzpicture}
\end{document}

>>>>

    \draw[edge,->] (0,-1) -- (0,-2.4);

    \node[vertex,fill=DodgerBlue,text=black] (s1) at (-2,-3) {$S_1^1$};
    \node[vertex,fill=DodgerBlue,text=black] (s2) at (2,-3) {$S_2^1$};
    \node[vertex,fill=DodgerBlue,text=black] (s3) at (0,-3) {$S_3^1$};
    \foreach \source/\dest in {3/2, 1/3}
    \draw[edge] (s\source) -- (s\dest);
  \end{tikzpicture}
\end{document}

### 2017-07-03 gtx_algorithm.tex
minor wording and layout adjustments

  Finally we prove that \gtx{} indeed returns a spanning tree of $G_0$. Specifically, we show that
  by starting from $G_K$ and going in reverse order, we build a spanning tree $T_t$ of every $G_t$,
  including eventually $G_0$. This process is illustrated on \autoref{fig:gtx_spanning_tree} and
  works as follow: when moving from $G_{t+1}$ to $G_t$, we expand the nodes of $G_{t+1}$ into star
  of nodes of $V_t$ and update all the outer edges we met so far to have their endpoints in $V_t$.
  These outer edges are exactly $T_{t+1}$ and by translating them into $E_t$ and adding the inner
  star edges of $G_t$, we build $T_t$. $G_K$ consists of a single node with no edges so we have
  trivially that $T_K$ = $G_K$. Then, assume we have a spanning tree $T_{t+1}$ of $G_{t+1}$, so that
  all nodes in $V_{t+1}$ are connected without cycle. Expanding each node of $V_{t+1}$ into a star
  of $V_t$ nodes ensures that the spanning property is preserved since each node of $V_t$ is covered
  by construction. Those stars are also without cycle by construction. The second step, translating
  the existing edges of $T_{t+1}$, only connects nodes belonging to different stars and because
  those edges were cycle free in $G_{t+1}$, this remains the case in $G_t$. Therefore we have build
  a spanning tree $T_t$ of $G_t$. By repeating this procedure, the \gtx{} algorithm eventually
  builds a spanning tree of $G_0$.

\end{proof}
\begin{figure*}[h]
  \centering
	\begin{subfigure}[b]{37mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g3_tikz.pdf}
		\caption{Spanning tree of $G_3$}\label{fig:gtx_spanning_tree_3}
	\end{subfigure}~
	\begin{subfigure}[b]{45mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g2_tikz.pdf}
		\caption{Spanning tree of $G_2$}\label{fig:gtx_spanning_tree_2}
	\end{subfigure}~
	\begin{subfigure}[b]{49mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g1_tikz.pdf}
		\caption{Spanning tree of $G_1$}\label{fig:gtx_spanning_tree_1}
	\end{subfigure}~
	\begin{subfigure}[b]{53mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g0_tikz.pdf}
		\caption{Spanning tree of $G_0$}\label{fig:gtx_spanning_tree_0}
	\end{subfigure}~
  \caption{Unfolding stars to recover spanning trees}
  \label{fig:gtx_spanning_tree}
\end{figure*}

@@

The action of \extractStar{} on $G_2$ is quite simple, because there is only one star that can be
created, so let say we choose $\startwo{1}$ as its center, with $\startwo{2}$ and $\startwo{3}$ as
peripheral nodes. This increases the eccentricity of their underlying $G_0$ nodes by $1$ (namely
nodes $16$ to $45$). Because there is only one star $\starthree{1}$ left, \collapseStar{} returns
$G_3$ showed in \autoref{fig:gtx_run3} and an empty list of outer edges, meaning that the inner loop
of \gtx{} is finished and we can go through every edges we chose between stars at every level to
recover the final spanning tree, showed with solid edges in \autoref{fig:gtx_eccentricity}
\vpageref{fig:gtx_eccentricity}. For completeness, we can also look at the edges which are not part
of the spanning tree and therefore contribute to the stretch of the tree. In that case the average
stretch is $7$, as showed in \autoref{tab:gtx_example_stretch}.

@@

A crucial quantity of the \gtx{} algorithm, both in terms of complexity and resulting stretch, is
the number of collapses $T$ needed before termination. While this is still elusive to express in the
general case, let us first look at some simple cases. For instance, a very sparse example of graph
is the line graph. While it is already a tree, let us look how the \gtx{} algorithm operates on it.
Say we have $n$ nodes and $m$ edges in that line. There a star is made at most of three
consecutive nodes. In the worst case, centers will be chosen such that we have a succession of three
and two nodes stars, as in \autoref{fig:gtx_line_graph}.%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.9\linewidth]{assets/tmp-code/line_graph.pdf}
  \caption{A line graph with stars in red}
  \label{fig:gtx_line_graph}
\end{marginfigure}
This will results in $\nicefrac{2n}{5}$ stars, which is less than half of $n$ and because in that
case $n=m+1$, \gtx{} will finish after $O(\log m)$ iterations. Note that a barbell graph (two
cliques connected by a line) would requires a number of iterations proportional to the length of that
central line, despite having many more edges. This suggests unsurprisingly that the diameter of the graph
could be a good parameter to quantify the number of iterations needed. Indeed, take a tree and consider
the length $p$ of its longest path from the root to a leaf. By a similar argument as the one used in
the line case, it seems \gtx{} will terminate after $O(\log p)$ iterations.

While all $G_t$ are connected, it might happen during the execution of \extractStar{} that%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_singleton_tikz.pdf}
  \caption{The formation of a singleton star}
  \label{fig:gtx_singleton}
\end{marginfigure}
the degree of a node $u$ drops to zero because all of its neighbors were claimed by the
periphery of previous stars. Such a node then forms a \emph{singleton star}, as illustrated in
\autoref{fig:gtx_singleton}, where after the creation of \starone{1} and \starone{2}, $c_3$ is the
single node of the third star \starone{3}.%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_noreduc_tikz.pdf}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_noreduc}
\end{marginfigure}
In the case no singleton are formed during one execution of \extractStar{}, the number of nodes is
reduced by at least a factor $2$ (because each star is made of at least two nodes). On the other hand, we can
construct a graph with singletons where the factor of reduction in number of nodes and edges can be
arbitrarily close to $1$. Consider \autoref{fig:gtx_noreduc}, where $c_1$ and $c_2$ have a degree
$p$ and their neighbors (in light gray) all have a degree $q$, which we can vary from $1$ to $p-1$.
At this step $t$, we thus have $|V_t| = 2(1+p)+pq$ and $|E_t| = 2(p+qp)$. Next we run \extractStar{}
and assume that we first extract the two stars centered in $c_1$ and $c_2$. This reduces the
effective degree of all the middle nodes (in dark gray) to $0$ and they thus form singletons, that
are all connected to the first two stars by \collapseStar{}. As a result, at step $t+1$ we have
$|V_{t+1}| = 2 + pq$ and $|E_{t+1}| = 2pq$, so that
\begin{equation*}
\frac{|V_{t+1}|}{|V_t|} = \frac{pq+2}{p(q+2) + 2} \qquad \text{and} \qquad
\frac{|E_{t+1}|}{|E_t|} = \frac{2pq}{2p(q+1)} = \frac{q}{q+1}
\end{equation*}
When we let $p$ goes to infinity, setting $q=1$ makes $\frac{|V_{t+1}|}{|V_t|}$ tend to
$\nicefrac{1}{3}$ and $\frac{|E_{t+1}|}{|E_t|}$ to $\nicefrac{1}{2}$ while setting $q=p-1$ makes
both $\frac{|V_{t+1}|}{|V_t|}$ and $\frac{|E_{t+1}|}{|E_t|}$ tend to $1$. Note however that in both
cases, this is not really a problem because at the next iteration, there will be only two stars (one
of them will be a singleton). Is it always like that? In other words, can we have a situation where
a large number of singletons is carried over several consecutive iterations?

>>>>

  Finally we prove that \gtx{} indeed returns a spanning tree of $G_0$. Specifically, we show that
  by starting from $G_K$ and going in reverse order, we build a spanning tree $T_t$ of every $G_t$,
  including eventually $G_0$. This process is illustrated on \autoref{fig:gtx_spanning_tree} and
  works as follow: when moving from $G_{t+1}$ to $G_t$, we expand the nodes of $G_{t+1}$ into star
  of nodes of $V_t$ and update all the outer edges we met so far to have their endpoints in $V_t$.
  These outer edges are exactly $T_{t+1}$ and by translating them into $E_t$ and adding the inner
  star edges of $G_t$, we build $T_t$. $G_K$ consists of a single node with no edges so we have
  trivially that $T_K$ = $G_K$. Then, assume we have a spanning tree $T_{t+1}$ of $G_{t+1}$, so that
  all nodes in $V_{t+1}$ are connected without cycle. Expanding each node of $V_{t+1}$ into a star
  of $V_t$ nodes ensures that the spanning property is preserved since each node of $V_t$ is covered
  by construction. Those stars are also without cycle by construction. The second step, translating
  the existing edges of $T_{t+1}$, only connects nodes belonging to different stars and because
  those edges were cycle free in $G_{t+1}$, this remains the case in $G_t$. Therefore we have build
  a spanning tree $T_t$ of $G_t$. By repeating this procedure, the \gtx{} algorithm eventually
  builds a spanning tree of $G_0$.
\end{proof}
\vspace{-2\baselineskip}
\begin{figure*}[h]
  \centering
	\begin{subfigure}[b]{37mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g3_tikz.pdf}
		\caption{Spanning tree of $G_3$}\label{fig:gtx_spanning_tree_3}
	\end{subfigure}~
	\begin{subfigure}[b]{45mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g2_tikz.pdf}
		\caption{Spanning tree of $G_2$}\label{fig:gtx_spanning_tree_2}
	\end{subfigure}~
	\begin{subfigure}[b]{49mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g1_tikz.pdf}
		\caption{Spanning tree of $G_1$}\label{fig:gtx_spanning_tree_1}
	\end{subfigure}~
	\begin{subfigure}[b]{53mm}
    \centering
		\includegraphics[width=\textwidth]{assets/tikz/gtx_st_g0_tikz.pdf}
		\caption{Spanning tree of $G_0$}\label{fig:gtx_spanning_tree_0}
	\end{subfigure}~
  \caption{Unfolding stars to recover spanning trees}
  \label{fig:gtx_spanning_tree}
\end{figure*}

@@

The action of \extractStar{} on $G_2$ is quite simple, because there is only one star that can be
created, so let say we choose $\startwo{1}$ as its center, with $\startwo{2}$ and $\startwo{3}$ as
peripheral nodes. This increases the eccentricity of their underlying $G_0$ nodes by $1$ (namely
nodes $16$ to $45$). Because there is only one star $\starthree{1}$ left, \collapseStar{} returns
$G_3$ showed in \autoref{fig:gtx_run3} and an empty list of outer edges, meaning that the inner loop
of \gtx{} is finished and we can go through every edges we chose between stars at every level to
recover the final spanning tree, a process illustrated in \autoref{fig:gtx_spanning_tree}
\vpageref{fig:gtx_spanning_tree}. For completeness, we can also look at the edges which are not part
of the spanning tree and therefore contribute to the stretch of the tree. In that case the average
stretch is $7$, as showed in \autoref{tab:gtx_example_stretch}.

@@

A crucial quantity of the \gtx{} algorithm, both in terms of complexity and resulting stretch, is
the number of collapses $K$ needed before termination. While this is still elusive to express in the
general case, let us first look at some simple cases. For instance, a very sparse example of graph
is the line graph. While it is already a tree, let us look how the \gtx{} algorithm operates on it.
Say we have $n$ nodes and $m$ edges in that line. Here, a star is made at most of three
consecutive nodes. In the worst case, centers will be chosen such that we have a succession of three
and two nodes stars, as in \autoref{fig:gtx_line_graph}.%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.9\linewidth]{assets/tikz/gtx_line_tikz.pdf}
  \caption{A line graph with stars in blue}
  \label{fig:gtx_line_graph}
\end{marginfigure}
This will results in $\nicefrac{2n}{5}$ stars, which is less than half of $n$ and because in that
case $n=m+1$, \gtx{} will finish after $O(\log m)$ iterations. Note that a barbell graph (two
cliques connected by a line) would requires a number of iterations logarithmic in the length of that
central line, despite having many more edges. This suggests unsurprisingly that the diameter of the graph
could be a good parameter to quantify the number of iterations needed. Indeed, take a tree and consider
the length $p$ of its longest path from the root to a leaf. By a similar argument as the one used in
the line case, it seems \gtx{} will terminate after $O(\log p)$ iterations.

While all $G_t$ are connected according to \autoref{lem:gtx_stay_connected}, it might happen during
the execution of \extractStar{} that%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_singleton_tikz.pdf}
  \caption{The formation of a singleton star}
  \label{fig:gtx_singleton}
\end{marginfigure}
the degree of a node $u$ drops to zero because all of its neighbors were claimed by the
periphery of previous stars. Such a node then forms a \emph{singleton star}, as illustrated in
\autoref{fig:gtx_singleton}, where after the creation of \starone{1} and \starone{2}, $c_3$ is the
single node of the third star \starone{3}.
In the case no singleton are formed during one execution of \extractStar{}, the number of nodes is
reduced by at least a factor $2$ (because each star is made of at least two nodes). On the other hand, we can
construct a graph with singletons where the factor of reduction in number of nodes and edges can be
arbitrarily close to $1$. Consider \autoref{fig:gtx_noreduc}, where $c_1$ and $c_2$ have a degree
$p$ and their neighbors (in light gray) all have a degree $q$, which we can vary from $1$ to $p-1$.
At this step $t$, we thus have $|V_t| = 2(1+p)+pq$ and $|E_t| = 2(p+qp)$. Next we run \extractStar{}
and assume that we first extract the two stars centered in $c_1$ and $c_2$. This reduces the
effective degree of all the middle nodes (in dark gray) to $0$ and they thus form singletons, that
are all connected to the first two stars by \collapseStar{}. As a result, at step $t+1$ we have%
\begin{marginfigure}
  \centering
  \includegraphics[width=0.95\linewidth]{assets/tikz/gtx_noreduc_tikz.pdf}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_noreduc}
\end{marginfigure}
$|V_{t+1}| = 2 + pq$ and $|E_{t+1}| = 2pq$, so that
\begin{equation*}
\frac{|V_{t+1}|}{|V_t|} = \frac{pq+2}{p(q+2) + 2} \qquad \text{and} \qquad
\frac{|E_{t+1}|}{|E_t|} = \frac{2pq}{2p(q+1)} = \frac{q}{q+1}
\end{equation*}
When we let $p$ goes to infinity, setting $q=1$ makes $\frac{|V_{t+1}|}{|V_t|}$ tend to
$\nicefrac{1}{3}$ and $\frac{|E_{t+1}|}{|E_t|}$ to $\nicefrac{1}{2}$ while setting $q=p-1$ makes
both $\frac{|V_{t+1}|}{|V_t|}$ and $\frac{|E_{t+1}|}{|E_t|}$ tend to $1$. Note however that in both
cases, this is not really a problem because at the next iteration, there will be only two stars (one
of them will be a singleton). Is it always like that? In other words, can we have a situation where
a large number of singletons is carried over several consecutive iterations?
